Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2023-05-30 Paper/PaperEdit.tex
Updates from Overleaf

\begin{abstract}
    Technological innovation has affected labor markets drastically over time. I add to our understanding of a new technology's affect on workers by investigating changes in physician behavior as a result of a major technology shift in U.S. health care, electronic health records (EHRs). I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital, and estimate average group time treatment effects on various labor market outcomes. I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs. Further, physicians exposed to EHRs see more patients due to the technology. These results point to unintended consequences of rapid technology implementation, but also indicate potential productivity gains for those who embrace the technology.
\end{abstract}

@@

\section{Introduction}
For decades, technological innovation has brought about changes to production in the economy through adjustments in wages, demand for labor, and other aspects of labor markets. Still, advancement of modern technology is rapid, and it is not clear how policymakers should be involved in the implementation of new technologies. This is particularly important in health care, where new technologies can impact patient outcomes. A recent technology roll-out that altered the US health care system dramatically was the electronic health record (EHR). Medical entities nationwide implemented EHRs rapidly from 2008-2015, with the percentage of hospitals with basic EHR capability increasing from 9 percent to 84 percent. Since many physicians working in hospitals were exposed to this technology in a concentrated time, this is an opportune setting to investigate the relationship between rapid technology implementation and individual labor market behaviors. In this paper, I investigate the effect of EHR adoption in hospitals on several behaviors of physicians working in hospitals, or hospitalists.

EHRs are computer systems that house digital medical records and sometimes have additional functionalities. EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech})\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The push for EHR use stemmed from a widespread expectation that EHRs would improve quality of health care while decreasing costs, the gold standard in health policy. For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).

@@

While hospitalists are ideal for to answer this question econometrically, they are also important in their own right. The Bureau of Labor Statistics estimates approximately 44,000 physicians employed as hospitalists in the US, and studies estimate that 75\% of hospitals employ hospitalists (\cite{bls}, \cite{healthecareers_2022}). Further, one third of Medicare admissions are seen by a hospitalist (\cite{messler2015history}). I construct a panel of hospitalists spanning from 2009-2017. Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify the hospital affiliation of each hospitalist. Then, I use the American Hospital Association (AHA) Survey to determine EHR adoption status, and thus connect hospitalists to EHR use. Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for information on the patients seen by each hospitalist. The independent variable of interest is binary, capturing exposure to an EHR.\footnote{While there are ample nuances to EHRs, I abstract away from these as I focus on the inital shock of a new technology.} I estimate group-time treatment effects of EHR exposure on the physician decisions defined above.

This paper contributes to our understanding of health information technology and its affects in health care. There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use. Despite a large number of case studies that find generally positive effects, i.e., improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), economic work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}). A number of case studies consider the effect of EHRs on productivity. One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EHRs being adopted in primary-care sites (\cite{Meyerhoefer}). There is still much to be learned about EHRs effects on health care, specifically related to labor supply (\cite{NBERw29218}).

@@

Two key assumptions underlie the main analysis. First, I assume that hospitalists are constrained to use the technology implemented in their work setting. Second, I assume the decision for hospitals to implement EHRs is exogenous to individuals. That is, a hospitalist employed by the hospital does not choose their own EHR exposure. My main findings are not sensitive to a plethora of robustness checks, and the results suggest that there was a surge of experienced hospitalists who left both clinical settings altogether and the hospitals where the EHR was deployed. Thus, EHR implementation may have intensified access to care issues for patients. This unintended consequence highlights the need for careful consideration of technologies and policies that affect work environment.

@@

\subsubsection{The Hospitalist Profession}

For ease of notation, I define a hospitalist in my data as any general practice physician who sees at least 70\% of their patients within a hospital setting. However, there are two types of physicians who can be categorized as hospitalists based on this definition: one being a formal hospitalist, and the other being an internist. Both formal hospitalists and internists are medical professionals who primarily focus on adult patient care, but they differ in their work settings. Internists are primary care physicians who provide comprehensive medical care to adult patients in either outpatient settings or hospitals, whereas formal hospitalists specialize in the care of hospitalized patients, and typically work exclusively in the hospital setting. While this distinction is potentially important for some of the decisions outlined below, I do not have the ability to distinguish between the two explicitly in the data. I will discuss this limitation in more detail when relevant.

@@

Age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. As a worker considers the retirement decision, they weigh their expected earnings with their expected costs, along with some preferences. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility partially depends on job satisfaction and personal life factors. Exogenous EHR implementation affects the decision to retire through job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Additionally, if EHRs cause physicians to substitute time away from patients and they receive less reimbursement, leisure becomes relatively cheaper, leading to retirement. Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), but this has not been addressed empirically.

Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards administrative roles, teaching, consulting, or hospital management. While it is unlikely that early career physicians are formally retiring, physicians of any age could be career switching. This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction. In a 2016 survey, 13\% of physicians stated that they were considering careers outside of clinical work because of burnout (\cite{physicians2016physicians}). As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch. Thus, the term exit for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards. A physician is said to exit in a given year only if it is the first year in which the physician has no future Medicare claims.

@@

For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of working in clinical settings. However, there are other ways hospitalists might change their behavior under drastic changes to their current working environment. For those who remain practicing, changing the physical place of work may allow for more control over which technologies they use. Some hospitalists work under contract full time for one medical practice where they have a consistent shift schedule (formal hospitalists), while others contract themselves out to multiple hospitals, offices, or both (internists). While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faced by each type.

First, I consider those who are in a relatively binding contract with the hospital. For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital. This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings. A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making it very costly to change work setting or excluding it from the choice set altogether. Since these hospitalists are established in their hospital, they work a significant amount of time there. When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings.

Alternatively, there may be hospitalists that work in multiple hospitals or work in both hospitals and offices throughout the sample period. If one hospital implements a costly (or beneficial) EHR, these workers can more easily shift all or some of their work towards (or away from) different facilities to allocate to the place where they gain the most job satisfaction. Theoretically, these types of hospitalists are much more likely to change their behavior as a result of a change in one work environment.

@@

There are still agents who are not induced to change their behavior on the basis of exit or work setting due to EHR implementation. A lack of response could be due to 4 factors: the hospitalist is bound to their current work environment, EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment. For hospitalists who make no physical changes to work environment after EHR exposure for any of these reasons, it is natural to assess whether EHRs affect productivity, as a main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.

Whether EHRs affect productivity in terms of patients seen is theoretically ambiguous. If they serve their purpose to decrease administrative burden, the physician would be freed up to see more patients. In this case, productivity would increase. However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity. Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the hospitalist. Billing activity refers to the number of claims filed per patient seen. EHRs may improve efficiency if they remove the need for repeated or excess testing. However, if having an EHR makes it easy for a physician to claim for additional items with the click of a button, the adverse incentive may lead physicians to claim for more items per patient.

@@

I construct a hospitalist-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labor market outcomes, and other relevant physician, hospital, and practice characteristics. I describe the different data sets used to construct the panel below, and I include a detailed outline of data and variable construction in Appendix \ref{app:data}.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office, emergency room, etc.). I only include physicians whose primary specialty as hospitalist\footnote{Physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting. I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year. I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate this threshold in Appendix \ref{sec:changes}). I drop hospitalists with low claim counts that do not accurately represent changes in behavior. I use this data to construct the dependent variables for my analysis. Generally, these include the decision to exit clinical work, switch to a different work setting, number of patients seen, and overall billing activity. I describe these outcome variables and their construction in detail in Section \ref{sec:outcome}.

@@

I aggregate the data to the physician level. Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015, as I do not observe whether they become treated in 2016 or 2017. That is, the data contains no units which are ``never-treated". Finally, I include a physician's graduation year from Physician Compare. I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship. The details of such thresholds are explained in Appendix \ref{app:data}. To ensure that my results  are not sensitive to various levels of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations. The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.

The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I refer to as exit. Exit is an indicator variable equal to 1 in the first year that a hospitalist sees no patients in any future year. I do not define retirement as cumulative (a treatment that turns on and remains on for all following year), but a one-time decision.\footnote{Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results. In Appendix \ref{sec:changes}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.} Next, I consider whether physicians exhibit behavior that suggests they have shifted work setting. I consider the fraction of patients a physician sees in an office-based/outpatient-based setting.. Then, I construct an indicator variable equal to one if a physician sees any positive number of patients in such a setting, which offers less variation but captures the extensive margin of working in an office. Finally, I consider outcome variables which explore productivity and efficiency, the number of patients seen and claims filed per patient.

@@

For each progression of decision making, I limit to physicians who did not change their behavior previously. That is, for office-based outcomes I exclude physicians who exit, and for productivity outcomes I exclude physicians who retire or change their work setting. I show summary statistics of relevant variables based on each sample in Table \ref{tab:sumstats}.

@@

\normalsize
I follow a maximum of 19,263 hospitalists over time, depending on the sample considered. Out of all hospitalists in the sample, 3\% (approximately 578) ever exit. A positive number of hospitalists are exposed to an EHR in each year, but the majority of exposure is concentrated towards the beginning of sample. For hospitalists in the office setting sample, 30\% of physicians have a positive number of patients seen in an office, and the average physician sees 9\% of their total patient volume in an office. For those who never change their work setting, the average hospitalist sees 325 patients and files 4 claims per patient. Consistent across all samples, hospitalists tend to work with 1.3 hospitals and 1.2 systems.

In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment over time. In 2009, 28\% of the sample of physicians were already exposed to an EHR. That is, three quarters of the sample had no affiliation with EHRs at the beginning of the sample period. Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by the end of the sample period. The darker line shows the underlying variation in this variable, hospitals implementing EHRs over time.

\begin{figure}[t]
\centering
\captionsetup{width=.45\linewidth}
    \caption{Treatment Variables Over Time}
    \includegraphics[scale=.4]{Objects/sum_stats_year.pdf}
    \label{fig:treatmentgraph}
\end{figure}

@@

I treat EHR implementation as an exogenous treatment variable, and I estimate average treatment effects of EHR exposure on the outcomes discussed. In this setting, treatment effects are heterogeneous across physicians who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt. Therefore, to avoid negative weighting issues this causes in a classical two way fixed effects event study specification, I instead estimate average treatment effects for a specific group $g$ at time $t$: 
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
where $G_g=1$ for those in group $g$. A group indicates all physicians treated, or first exposed to an EHR, in the same year. To estimate the heterogeneous treatment effects, I employ the doubly robust estimator established in \citeauthor{sant2020doubly} (\citeyear{sant2020doubly}). Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}. I also present classical two way fixed effects results with physician and year fixed effects in Appendix \ref{app:estimators}.

I aggregate the group time estimates over groups to a more familiar event study plot. For select outcomes, I present the original pre-aggregation estimates to show that results are not driven by a particular year of treatment.  Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each results plot. Simultaneous confidence bands, calculated using bootstrapped standard errors, are also presented.

@@

There are several assumptions necessary to identify the parameters of interest, $ATT(g,t)$. First, I assume that treatment is not reversed. That is, once a physician is exposed to an EHR, they cannot be un-exposed. For physicians that do not switch hospitals, this assumption is supported by the institutional details of hospital EHR implementation. An EHR is a costly technology and requires a significant amount of collaboration to implement. A hospital does not have incentive to un-implement an EHR once it is implemented. They may add features or switch vendors, but do not reverse EHR use. However, when considering the outcomes of patient count and billing activity, it could be that a physician moves from a hospital that uses an EHR to a hospital that does not, affecting the number of claims observed in the data. Thus, I limit the sample of physicians for these outcomes to those who remain with the same set of hospitals throughout the entire period. I discuss this limitation to the data further in Section \ref{sec:patientcount}.

Second, I assume physicians do not anticipate EHR exposure prior to occurrence. This may be a concern if physicians learn that the system will be implemented and change their behavior beforhand. Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}). Therefore, I proceed in the main specification assuming no anticipation. However, I explore and present results for a one year anticipation period in the specification charts in Appendix \ref{sec:changes}, and I find that accounting for one year of anticipation does not drastically change my findings.

>>>>

\begin{abstract}
    Technology adoption in health care is of great importance to policy makers as some technologies are expected to radically improve the U.S. health care system. However, some new technologies can be burdensome for physicians and lead to frustration. I add to our understanding of a new technology's affect on workers by investigating changes in physician behavior as a result of a major technology shift in U.S. hospitals, electronic health records (EHRs). I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital (hospitalists), and estimate average group time treatment effects on various labor market outcomes. I find that hospitalists are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs. Further, those exposed to EHRs see more patients due to the technology. These results point to unintended consequences of rapid technology implementation, but also indicate potential productivity gains for those who embrace the technology.
\end{abstract}

@@

\section{Introduction}
For decades technological innovation has brought about changes to production in the economy through adjustments in wages, demand for labor, and other aspects of labor markets. Still, advancement of modern technology is rapid, and it is not clear how policymakers should be involved in the implementation of new technologies. This is particularly important in health care, where new technologies can impact patient outcomes. A recent technology roll-out that altered the US health care system dramatically was the electronic health record (EHR). Medical entities nationwide implemented EHRs rapidly from 2008-2015, with the percentage of hospitals with basic EHR capability increasing from 9 percent to 84 percent. Since many physicians working in hospitals were exposed to this technology in a concentrated time, this is an opportune setting to investigate the relationship between rapid technology implementation and individual labor market behaviors. In this paper, I investigate the effect of EHR adoption in hospitals on several behaviors of physicians working in hospitals.

EHRs are computer systems that house digital medical records and sometimes have additional functionalities. EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech}).\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})} President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The push for EHR use stemmed from a widespread expectation that EHRs would improve quality of health care while decreasing costs, the gold standard in health policy. For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).

@@

While hospitalists create an ideal setting to answer this question econometrically, they are also important in their own right. The Bureau of Labor Statistics estimates approximately 44,000 physicians employed as hospitalists in the US, and studies estimate that 75\% of hospitals employ hospitalists (\cite{bls}, \cite{healthecareers_2022}). Further, one third of Medicare admissions are seen by a hospitalist (\cite{messler2015history}). I construct a panel of hospitalists spanning from 2009-2017. Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify the hospital affiliation of each hospitalist. Then, I use the American Hospital Association (AHA) Survey to determine EHR adoption status, and thus connect hospitalists to EHR exposure. Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for information on the patients seen by each hospitalist. The independent variable of interest is binary, capturing exposure to an EHR.\footnote{While there are ample nuances to EHRs, I abstract away from these as I focus on the inital shock of a new technology.} I estimate group-time treatment effects of EHR exposure on the physician decisions defined above.

This paper contributes to our understanding of the effects of health information technology. There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use. Despite a large number of case studies that find generally positive effects, i.e., improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), economic work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}). A number of case studies consider the effect of EHRs on productivity. One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EHRs being adopted in primary-care sites (\cite{Meyerhoefer}). There is still much to be learned about EHRs effects on health care, specifically related to labor supply (\cite{NBERw29218}).

@@

Two key assumptions underlie the main analysis. First, I assume that hospitalists are constrained to use the technology implemented in their work setting. Second, I assume the decision for hospitals to implement EHRs is exogenous to individuals. That is, a hospitalist employed by the hospital does not choose their own EHR exposure. My main findings are not sensitive to a plethora of robustness checks, and the results suggest that there was a surge of experienced hospitalists who left both clinical settings altogether and the hospitals where the EHR was deployed. Thus, EHR implementation may have intensified access to care issues for patients. This unintended consequence highlights the need for careful consideration of technologies and policies that affect work environment.  \textcolor{red}{reword the implications and check the numbers are correct}

@@

For ease of notation, I define a hospitalist in my data as any general practice physician who sees at least 70\% of their patients within a hospital setting. However, there are two types of physicians who can be categorized as hospitalists based on this definition: one being a formal hospitalist, and the other being an internist. Both formal hospitalists and internists are medical professionals who primarily focus on general patient care, but they potentially differ in their work settings. Internists are primary care physicians who provide comprehensive medical care to adult patients in either outpatient settings or hospitals, whereas formal hospitalists specialize in the care of hospitalized patients, and typically work exclusively in the hospital setting. While this distinction is potentially important for some of the decisions outlined below, the data do not explicitely distinguish between the two. I will discuss this limitation in more detail when relevant.

@@

Age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. As a worker considers the retirement decision, they weigh their expected earnings with their expected costs, along with some preferences. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility partially depends on job satisfaction and personal life factors. Exogenous EHR implementation affects the decision to retire through job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), but this has not been addressed empirically.

Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards administrative roles, teaching, consulting, or hospital management. While it is unlikely that early career physicians are formally retiring, physicians of any age could be career switching. This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction. In a 2016 survey, 13\% of physicians stated that they were considering careers outside of clinical work because of burnout (\cite{physicians2016physicians}). As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch. For the purpose of this paper, the term exit indicates no longer seeing patients, and makes no assertion about what the physician does afterwards.

@@

Most physicians are likely not on the margin in their decision to exit clinical work, so EHRs will not impose a large enough cost (or benefit) to impose exit. However, there are other ways hospitalists might change their behavior under drastic changes to their current working environment. For those who remain practicing, changing the physical place of work may allow for more control over which technologies they use. Specifically, if EHRs change job satisfaction of working in the hospital, a hospitalist may decide to shift more of their working hours towards or away from the hospital. Not all physicians working as hospitalists have a binding contract with their associated hospital. In fact, a 2013 survey estimated that 36\% of hospitalists also worked outside of the hospital in primary care, long-term care, or hospice settings (\cite{Today}). Many physicians trained as internists choose to contract their work out to both hospitals and outpatient settings. The fraction of internists who work as hospitalists increased from 25\% to 40\% from 2008 to 2018 (\cite{gray2022evolving}).

The sample of physicians I consider include both hospitalists with formal contracts and internists who contract a majority of their time with the hospital, but could also work in other settings as well. If a hospital implements a costly EHR, a hospitalist may choose to shift more of their time towards outpatient settings or initiate a contract with outpatient settings to spend less time in the setting where job satisfaction has decreased.

@@

For those not induced to change their behavior by exiting or changing work settings, it is natural to assess whether EHRs impacted productivity. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden. This question is empirically ambiguous since EHRs could add daily tasks that make hospitalists less productive, increase efficiency, or some combination of the two where short term productivity is negatively impacted but long term productivity is positively impacted once physicians adjust and learn the new technology. I measure productivity in terms of the amount of patients a hospitalist is able to see.

Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the hospitalist. Billing activity refers to the number of claims filed per patient seen. There are many reported reasons that physicians tend to over-treat patients, including defensive medicine, patient request, difficulty accessing patient records, and pressure from their institution (\cite{lyu2017overtreatment}). EHRs have the potential to change some of these incentives. As more hospitals adopt EHRs, patient information should be more easily accessible and reduce the amount of repeated tests a physician needs to run, leading to a decrease in claims per patient. However, institutions have power to customize EHRs and train their hospitalists to increase fee-for-service revenue by increasing claims per patient.

@@

I construct a hospitalist-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labor market outcomes, and other relevant physician and hospital characteristics. I describe the different data sets used to construct the panel below, and I include a detailed outline of data and variable construction in Appendix \ref{app:data}.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office/outpatient, emergency room, etc.). I only include physicians classified as hospitalists\footnote{Physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting. I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year. I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate this threshold in Appendix \ref{sec:changes}). I use this data to construct the dependent variables for my analysis. Generally, these include the decision to exit clinical work, switch to a different work setting, number of patients seen, and billing activity.

@@

I aggregate the data to the physician level. I drop physicians who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated. The results are robust to this specification choice.} That is, the data contains no units which are ``never-treated". Finally, I include a physician's graduation year from Physician Compare (provided by CMS). I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship. The details of such thresholds are explained in Appendix \ref{app:data}. To ensure that my results  are not sensitive to various levels of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations. The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.

The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I refer to as exit. Exit is an indicator variable equal to 1 in the first year that a hospitalist sees no patients in any future year and zero in every other year. This captures a one time decision to stop seeing patients, not a cumulative decision in all following years. The next variables I construct capture whether physicians change work setting. I consider the fraction of patients a physician sees in an office-based/outpatient setting, and I construct an indicator variable equal to one if a physician sees any positive number of patients in such a setting. Finally, I consider outcome variables which explore productivity and efficiency, the number of patients seen and claims filed per patient.

@@

For each progression of decision making, I limit to physicians who did not change their behavior previously. That is, for office-based outcomes I exclude physicians who exit, and for productivity outcomes I exclude physicians who exit or change their work setting. I show summary statistics of relevant variables based on each sample in Table \ref{tab:sumstats}.

@@

\normalsize
I follow a maximum of 19,263 hospitalists over time, depending on the sample considered. Out of all hospitalists in the sample, 3\% (approximately 578) stop seeing patients at some point. A positive number of hospitalists are exposed to an EHR in each year, but the majority of exposure is concentrated towards early years. For hospitalists in the office setting sample (those who do not exit), 30\% of physicians have a positive number of patients seen in an office, and the average physician sees 9\% of their total patient volume in an office. There are 12,732 hospitalists who never exit or change their work setting at any point in the sample. These hospitalists see 325 patients and file 4 claims per patient on average. Consistent across all samples, hospitalists tend to work with 1.3 hospitals and 1.2 systems.

In Figure \ref{fig:treatmentgraph}, I present a graph of the variation in treatment over time and the underlying hospital adoption that drives this treatment variable. Hospitalists are exposed the first time one of their hospital's adopts an EHR. Thus, I show the percent of hospitals adopting as well as the percent of hospitalists exposed to the technology. In 2009, 41\% of all hospitalists were already exposed to an EHR. That is, half of the sample had no affiliation with EHRs at the beginning of the sample period. Since I drop any physicians who do not have exposure to an EHR by 2015, 100\% of the sample is exposed by the end of the sample period. In 2009, 47\% of hospitals have adopted an EHR, and this percentage grows steadily over time, driving the increase in exposed hospitalists.

\begin{figure}[t]
\centering
\captionsetup{width=.45\linewidth}
    \caption{Treatment Variables Over Time}
    \includegraphics[scale=.6]{Objects/sum_stats_year.pdf}
    \label{fig:treatmentgraph}
\end{figure}

@@

I estimate average treatment effects of EHR exposure on the outcomes discussed. In this setting, treatment effects may be heterogeneous across hospitalists who are exposed in different years since the underlying characteristics of hospitals who adopt in different years may be correlated with the decision to adopt. Therefore, to avoid negative weighting issues this causes in a classical two way fixed effects event study specification, I instead estimate average treatment effects for a specific group $g$ at time $t$: 
$$ATT(g,t)=\mathbbm{E}[Y_t(g)-Y_t(0)|G_g=1],$$
where $G_g=1$ for those in group $g$. A group indicates all hospitalists treated, or first exposed to an EHR, in the same year. I employ the doubly robust estimator established in \citeauthor{sant2020doubly} (\citeyear{sant2020doubly}). Other estimators that similarly address the concerns yield similar results, presented in Appendix \ref{app:estimators}. I also present classical two way fixed effects results with physician and year fixed effects in Appendix \ref{app:estimators}.

I aggregate the group time estimates over groups to a more familiar event study plot with simultaneous confidence bands calculated using bootstrapped standard errors. Further, a weighted average of estimates is taken to provide a single $ATT$ value for each outcome; these values are presented in the notes of each results plot.

@@

There are several assumptions necessary to identify the parameters of interest, $ATT(g,t)$. First, I assume that treatment is not reversed. That is, once a hospitalist is exposed to an EHR, they cannot be un-exposed. For hospitalists that do not switch hospitals, this assumption is supported by the institutional details of hospital EHR implementation. An EHR is a costly technology and requires a significant amount of collaboration to implement, so a hospital does not have incentive to un-implement an EHR. They may add features or switch vendors, but do not reverse EHR use. Any hospitalists included in the productivity sample are those which did not switch work settings, satisfying this assumption.

Second, I assume hospitalists do not anticipate EHR exposure prior to occurrence. If they learn that the system will be implemented and change their behavior beforehand, the estimates may be biased. Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}). Therefore, I proceed in the main specification assuming no anticipation. However, I explore and present results for a one year anticipation period in the specification charts in Appendix \ref{sec:changes}, and I find that accounting for one year of anticipation does not drastically change the findings.

As is usual in a difference-in-differences framework, I assume a version of parallel trends based on not-yet-treated units. I assume that, conditional on years of experience, average outcomes for those treated in group $g$ would have followed a parallel trend as those in groups treated in later periods. While this assumption cannot be tested directly, I present p-values for a Wald test of pre-trends. Further, I investigate pre-trends extensively in Appendix \ref{sec:pretrends} by allowing for specified violations in the parallel trends assumption (\cite{rambachan2019honest}). I find that most outcomes are very robust to violations in the parallel trends assumption.

To interpret these estimators as the causal effect of EHR exposure on various labor market outcomes, there are additional institutional assumptions necessary. First, I assume that when a hospitalist works in an electronic record utilizing-hospital, they utilize the electronic health record system fully. That is, the hospitalist does not continue practicing without using the EHR. While some hospitalists may create work-arounds to ease the burden of EHR use, it is inevitable that they will need to interact with the technology in some capacity. Second, I assume that a physician's exposure to EHRs is exogenous. That is, the physician does not influence the hospital's decision to implement an EHR.
